---
title: "Methods"
author: Steven Bachman
date: 29th October 2018
output: md_document
---

This document describes methods used in the paper **'Progress, challenges and opportunities for plant Red Listing'**. Each numbered section relates to numbered sections in the manuscript. See section [04_docs](https://github.com/stevenpbachman/RedList_Challenges/tree/master/02_docs) for the manuscript.

### 1. Introduction 

```{r echo=FALSE}
# RL key needed for download. ***change this to your own key- get from iucn.org 
rlkey = "cf3cf316d5dd72945456d42a633ac92357f6b8039bf1d512cc58e26d6169a8f3"
#rlkey <- "9bb4facb6d23f48efbf424bb05c0c1ef1cf6f468393bc745d42179ac4aca5fee" # IUCN key
```

*Total number of taxa on the Red List:*
```{r}
#install.packages("rredlist")
library(rredlist)

# get the total number of taxa
rl_count = rredlist::rl_sp_count(key = rlkey)
rl_count = rl_count$count

# note the version of the Red List
rl_ver = rredlist::rl_version(key = rlkey)
```

**Number of taxa on the Red List = `r rl_count`**


**Red List version = `r rl_ver`**




*Total number of species on the Red List:*
```{r}
# this function will get basic data from the Red List
rl_all_IDs = function(rlkey){
  
  # get page with 10,000   
  i=1
  testcount = 1
  rl = rl_sp(0, key = rlkey)
  rl.all = rl$result
  
  # use 'while' to loop through pages until there are no more pages
  while ((testcount)>'0'){
    test = rl_sp(i, key = rlkey)
    testcount = test$count
    rl.all = rbind(rl.all,test$result)
    i = i+1
  }
return(rl.all)  

}
```

Run the rl_all_IDs function to get the full red list of taxa. 
Subset to remove infra rank and populations to get all species

```{r}
# run rl_all_IDs to get dataframe of all taxa on Red List 
all_taxa = rl_all_IDs(rlkey)
```
     
```{r}     
# subset to remove subspecies and infra ranks
all_RL_species = subset(all_taxa, is.na(all_taxa$infra_rank) & is.na(all_taxa$population))

# Get total number of 'species' on Red List
rl_row = nrow(all_RL_species)
```

**Number of species on the Red List = `r rl_row `**





### 1.1 Gaps in Red List coverage – why the missing species matter
*Percentage of described vertebrate species that have been Red Listed:*

Estimated number of described species of vertebrates, invertebrates, plants and fungi. 
Source: http://cmsdocs.s3.amazonaws.com/summarystats/2018-1_Summary_Stats_Page_Documents/2018_1_RL_Stats_Table_1.pdf

```{r}
# subset all species to get only vertebrates
all.verts = subset(all_RL_species, all_RL_species$phylum_name == "CHORDATA")

# estimated number of described species of VERTEBRATES
described.verts = 69276

# proportion vert species described that have been Red Listed
prop.verts.RL = round((nrow(all.verts)/described.verts)*100, digits = 2)
```

**Proportion of vertebrate species described that have been Red Listed = `r prop.verts.RL `%**



*Percentage of described invertebrates, plants and fungi species that have been Red Listed:*

```{r}
# subset all species to get only plants
all.plants = subset(all_RL_species, all_RL_species$kingdom_name == "PLANTAE") #KINGDOM = PLANTAE
all.plants.count = nrow(all.plants)
```

**Number of plant species Red Listed = `r all.plants.count `**



```{r}
# fungi and protists covered in two kingdoms so use 'which' and pipe symbol to combine
all.fungi = all_RL_species[ which(all_RL_species$kingdom_name == "FUNGI" | all_RL_species$kingdom_name == "CHROMISTA"),]
all.fungi.count = nrow(all.fungi)
```

**Number of fungi species Red Listed= `r all.fungi.count `**



```{r}
# inverts is any animalia that isn't phylum CHORDATA so use kingdom == ANIMALIA & phylum != CHORDATA
all.inverts = all_RL_species[ which(all_RL_species$kingdom_name == "ANIMALIA" & all_RL_species$phylum_name != "CHORDATA"),]
all.inverts.count = nrow(all.inverts)
```

**Number of invertebrate species Red Listed = `r all.inverts.count `**



```{r}
# combine all the non-vertebrates 
all.RL.nonverts =  all.plants.count + all.fungi.count + all.inverts.count
```



**Number of non-vertebrate species Red Listed = `r all.RL.nonverts `**


Estimated number of described species of invertebrates, plants and fungi. 
Source: http://cmsdocs.s3.amazonaws.com/summarystats/2018-1_Summary_Stats_Page_Documents/2018_1_RL_Stats_Table_1.pdf

For vascular plants source:  
Lughadha, E.M., Govaerts, R., Belayaeva, I., Black, N., Lindon, H., Allkin, R., Magill, R.E., Nicholson, N., 2016. Counting counts: revised estimates of numbers of accepted species of flowering plants, seed plants, vascular plants and land plants with a review of other recent estimates. Phytotaxa 272, 82–88.
https://www.biotaxa.org/Phytotaxa/article/view/phytotaxa.272.1.5

```{r}
described.vasc.plants.niclughadha2016 = 383671
described.fungi = 52280
described.inverts = 1305250
described.nonvert = described.vasc.plants.niclughadha2016 + described.fungi + described.inverts

# proportion described invertebrates, plants and fungi species that have been Red Listed
prop.nonverts.RL = all.RL.nonverts/described.nonvert*100
```

**Proportion of invertebrates described that have been Red Listed = `r round((prop.nonverts.RL), digits = 2) `%**



### 1.2 Growing the Red List - vascular plants as a case study
Filter on vascular plants only and compare with best estimate for total number of vascular plants i.e. Nic Lughadha et al 2016
```{r}
vasc_plants = subset(all.plants, all.plants$phylum_name == "TRACHEOPHYTA") #PHYLUM = TRACHEOPHYTA (vascular plants)
prop_vasc = nrow(vasc_plants)/described.vasc.plants.niclughadha2016*100
```

**Proportion of vascular plants that have been Red Listed = `r round((prop_vasc), digits = 2)`**



### 2.3 Batch assessment upload with ‘SIS Connect’
Percentage of assessments published on the Red List via SIS Connect
Craig Hilton-Taylor, pers. comm (10th August 2018)

```{r}
sis_connect_pub = 509
sis_connect_pre_pub = 915
```
**Assessments published via SIS Connect = `r sis_connect_pub`**


**Assessments in SIS Connect pipeline = `r sis_connect_pre_pub`**



### 2.4 Inclusion of assessments in languages other than English

List of mega diverse countries derived from this source:
http://www.biodiversitya-z.org/content/megadiverse-countries.pdf

```{r}
# read in table of mega diverse countries
mega_diverse_countries <- read.csv("01_data/mega_diverse countries.csv")

# get the number of megadiverse countries that speak Spanish, Portuguese or French as main language
country.count = nrow(subset(mega_diverse_countries, mega_diverse_countries$Span.Port.or.French == "Yes"))
```
**Number of mega diverse countries that Spanish, Portuguese or French = `r country.count`**



```{r}
# Red list assessments in other languages
brazil_2016_prt = 20
haiti_2018_1_frnch = 38
```



### 2.5 Spatial tools support Red List automation
Craig Hilton-Taylor, pers. comm (10th August 2018)
```{r}
spatial_tools_citations = 1939
```


**Number of publications citing spatial tools =  `r spatial_tools_citations`**




### 2.6 Linking new  species and Red List assessment publications 
To get the average number of new taxa described every year, data were downloaded from IPNI Beta site: https://beta.ipni.org/

A query was made for each year, with a filter on 'Specific' taxa (species). A csv for each year from 1999 - 2017 inclusive was downloaded. See [01_data](https://github.com/stevenpbachman/RedList_Challenges/tree/master/01_data/IPNI_downloads_25_07_2018) for raw data.


```{r}
# read in raw files. Thanks to Hayward Godwin: https://www.r-bloggers.com/merge-all-files-in-a-directory-using-r-into-a-single-dataframe/

# get list of files, note full.names = true to get full path
file_list = list.files("01_data/IPNI_downloads_25_07_2018/", full.names = TRUE)

#create first dataset
IPNI_raw <- read.csv("01_data/IPNI_downloads_25_07_2018/citations_1999.txt")

# make blank DF with structure
IPNI_raw = IPNI_raw[0,]

for (file in file_list){
       
# if the merged dataset does exist, append to it
  if (exists("IPNI_raw")){
    temp_dataset <-read.csv(file)
    IPNI_raw = rbind(IPNI_raw, temp_dataset)
    rm(temp_dataset)
  }
}
```

```{r}
# save this down
path = getwd()
respath = paste0(path,"/04_outputs/IPNI_raw.csv")
write.table(IPNI_raw, respath,row.names = FALSE, na="", sep = ",")
```

Subset the IPNI data to just tax.nov.
```{r}
tax.nov = subset(IPNI_raw, IPNI_raw$citation.type == "tax. nov.")
```

Summarise number of tax.nov by year
```{r echo=FALSE}
# load library
library(tidyverse)
```


```{r}
tax.nov.acc = subset(tax.nov, name.status.type == "")
tax.nov.year = tax.nov.acc %>% dplyr::group_by(published) %>% dplyr::count(published)
```

Plot number of newly described species per year
```{r}
colnames(tax.nov.year)[which(names(tax.nov.year) == "published")] = "Year"
colnames(tax.nov.year)[which(names(tax.nov.year) == "n")] = "Count"

library(ggplot2)
plot = ggplot(data = tax.nov.year) + 
  geom_col(aes(x = Year, y = Count))
plot
```

```{r}
# get average number of described species between 1999 and 2017
tax.nov.year.mean = round((mean(tax.nov.year$Count)),0)
```


**The description of new plant species occurs at a fairly consistent rate, with a mean of `r tax.nov.year.mean` per year (1999 – 2017)**


Kew Bulletin example - how many newly described taxa published in Kew Bulletin from 2003–2017 are currently on the Red List?

```{r}
# get Kew Bull species 2003 - 2017
#create first dataset
KewBull_raw <- read.csv("01_data/IPNI_Kew_Bulletin_IPNI_downloads_25_07_2018/citations_KewBull_2003-2017.txt")

# summarise to just tax.nov
KewBull_tax.nov = subset(KewBull_raw, KewBull_raw$citation.type == "tax. nov.")
nrow(KewBull_tax.nov)

# remove invalid
KewBull_tax.nov = subset(KewBull_tax.nov, name.status.type == "")

# change column names to match
colnames(KewBull_tax.nov)[which(names(KewBull_tax.nov) == "scientific.name")] = "scientific_name"

# now link to the red list to see how many have been assessed
KB_RL_joined = merge(KewBull_tax.nov, all.plants, by = "scientific_name" )

# Get totals
nrow(KB_RL_joined)
nrow(KB_RL_joined)/nrow(KewBull_tax.nov)*100
```

**Of the 1,234 newly described taxa published in Kew Bulletin from 2003–2017, only `r nrow(KewBull_tax.nov)`, `r round((nrow(KB_RL_joined)/nrow(KewBull_tax.nov)*100),digits = 2)`% are currently on the Red List**


How much time has elapsed between formal description of a species and publication of an assessment for that species on the Red List? Species described and then assessed within 5 years were labelled as ‘New’, and species that were assessed more than 5 years since they were described were labelled ‘Old’. 

1. Use [POWO](http://www.plantsoftheworldonline.org/) API to assign IPNI IDs to Red List plant names.
2. Then use [OpenRefine](http://openrefine.org/) and IPNI ID to get date/year that plant name description was published
3. Query Red List to get assessment year (this doesn't come with the summary data)
4. Compare Red List assessment year with name description publication year - plot results.


1. Use [POWO](http://www.plantsoftheworldonline.org/) API to assign IPNI IDs to Red List plant names.
```{r eval=FALSE}
### Query Red List binomials against POWO to see taxonomic status according to POWO
### POWO = Plants of the World Online

library(jsonlite)
library(tidyverse)
library(plyr)

check.accepted.POWO = function(binomial) {
  
  #binomial = "Solanum polygamum"
  
  # use ID full name to search API  
  full_url =  paste("http://plantsoftheworldonline.org/api/1/search?q=names:", binomial, sep = "")
  
  # get raw json data
  raw.data <- readLines(full_url, warn = "F", encoding = "UTF-8")
  
  # organise
  rd <- fromJSON(raw.data)
  
  if (length(rd$results) == 0) {
    # add new column to results and call it warning
    
    accepted = "NA"          
    author = ""  
    kingdom = ""  
    name = ""  
    rank  = "" 
    #synonymOf = ""
    base_url = ""    
    IPNI_ID = ""
    search_name = binomial  
    results = data.frame(accepted,author, kingdom, name, rank, base_url, IPNI_ID,search_name)
    
  } else {
    
    # make data frame
    results = as.data.frame(rd$results)
    
    # add original search term
    results$search_name = binomial
    
    # returns null because there is more than one result - come back to this later and print out extra rows
    if (nrow(results) >1) {
      
      accepted = "FALSE"          
      author = ""  
      kingdom = ""  
      name = ""  
      rank  = "" 
      #synonymOf = ""
      base_url = ""    
      IPNI_ID = ""
      search_name = binomial  
      results = data.frame(accepted,author, kingdom, name, rank, base_url, IPNI_ID,search_name)
      
    } else {
      
      # PROBLEM HERE - var with no author - replace with blank field?
      if (!"author" %in% colnames(results)) {
        results$author = NA
      }
      
      # only include these fields - you don't want synonym of
      results = subset(results, select=c(accepted, author, kingdom,name,rank, url, search_name))
      
      # split url to get url and IPNI_ID
      results = results %>% separate(url, into = c("base_url", "IPNI_ID"), sep = "names:")
      
      # take out any results where it matched on something that wasn't species 
      #results = subset(results, rank == "Species") 
      
      # check if 
      if (nrow(results) < 1) {
        # add new column to results and call it warning
        
        accepted = "NA"          
        author = ""  
        kingdom = ""  
        name = ""  
        rank  = "" 
        #synonymOf = ""
        base_url = ""    
        IPNI_ID = ""
        search_name = binomial  
        results = data.frame(accepted,author, kingdom, name, rank, base_url, IPNI_ID,search_name)
        
      }
      # make data frame
      results = as.data.frame(results)
    }
    
    return(results)
    
  }
}

#check if names are accepted on POWO - query on the binomial field - change this for exact column name rather than col number?
checkACC_apply = lapply(all.plants[,8],check.accepted.POWO)
accepted.result = do.call(rbind.fill, checkACC_apply)

# reduce to just the rows with IPNI IDs - convert blanks to NA first
accepted.result$IPNI_ID[accepted.result$IPNI_ID==""] = NA

# omit NAs
all.plants.IPNI = accepted.result %>% na.omit()

# reduce to non null results
all.plants.IPNI = subset(all.plants.IPNI, rank =="Species") 

# keep accepted red_list in memory, but save as hard copy for later use if necessary
path = getwd()
respath = paste0(path,"/all.plants.IPNI.csv")
write.table(all.plants.IPNI, respath,row.names = FALSE, na="", sep = ",")
```

2. Then use [OpenRefine](http://openrefine.org/) and IPNI ID to get date/year that plant name description was published

Now that we have IPNI IDs for the Red List assessments for plants, we need to pull out the date/year of 
publication so we can compare assessment date vs publication date. 

Use [OpenRefine](http://openrefine.org/) to access structured IPNI data.
Thanks to [Nicky Nicholson](https://github.com/nickynicolson) for method.

1.	Call IPNI to get structured data for each record:
⋅⋅*	Click dropdown next to the IPNI id, select “Edit column” ->  “Add column by fetching URLs”
⋅⋅*	Build the URL by entering the following into the expression box:
'http://ipni.org/urn:lsid:ipni.org:names:'+value
⋅⋅*	Give your new column a title (e.g. “IPNI LSID data”) and a throttle delay of 1000 milliseconds, click OK
You should now have a new column with RDF-XML formatted data – these cells will be massive but you can drop (or hide) the column after the next step is completed
2.	Extract the publication year from the RDF XML:
⋅⋅*	Click the dropdown next to the “IPNI LSID data” column, select “Edit column” -> “Add column based on this column”
⋅⋅*	Select the year tag in the RDF XML and extract its contents: type in:
value.parseHtml().select("tn|year")[0].htmlText()
⋅⋅*	Give your new column a title (‘publication year’) and click OK

Save as [RedList2018_1_powo_species_IPNI_LSID](https://github.com/stevenpbachman/RedList_Challenges/blob/master/01_data/RedList2018_1_powo_species_IPNI_LSID.csv)

3. Query Red List to get assessment year (this doesn't come with the summary data)
```{r eval = FALSE}
#get red list year
# import the names list with IPNI ID and year
path = getwd()
all_plants_IPNI <- read_csv(paste0(path,"/01_data/all.plants.IPNI.csv"))

# use the binomial to get the Red List extra data
IPNI_year = all_plants_IPNI
rl_names = IPNI_year[,4]

# load the red list library
library(rredlist)

#run this function to get the extra red list data
rl = function(x, key){
  sp = rl_search(name = x, key = key)
  sp = sp$result
  return(sp)
}

# use apply to run through all in the list
t = proc.time()
test_apply = apply(rl_names, 1, rl, key = rlkey)
rl_df = as.data.frame(do.call(rbind,test_apply))
proc.time()- t

# now join back with all_plants_IPNI
# change column name so that you can merge with full red list data
colnames(all_plants_IPNI)[which(names(all_plants_IPNI) == "search_name")] = "scientific_name"
rl_IPNI = merge(all_plants_IPNI, rl_df, by = "scientific_name")
```


4. Compare Red List assessment year with name description publication year - plot results.

```{r eval = FALSE}
path = getwd()
RedList2018_1_powo_species_IPNI_LSID <- read.csv(paste0(path,"/01_data/RedList2018_1_powo_species_IPNI_LSID.csv"))
rl_IPNI_year = RedList2018_1_powo_species_IPNI_LSID
rl_IPNI_year = subset(rl_IPNI_year, select = -IPNI_LSID)
colnames(rl_IPNI_year)[which(names(rl_IPNI_year) == "search_name")] = "scientific_name"

# merge IPNI year and Red List 
merged_taxnov_RL = merge(rl_IPNI, rl_IPNI_year, by = "scientific_name")

# now take the described year away from the published year
merged_taxnov_RL_diff = data.frame(diff = merged_taxnov_RL$published_year - merged_taxnov_RL$Year)
diff = cbind(merged_taxnov_RL_diff, merged_taxnov_RL)

# now add column for New (5 years or less) and Old (>5 years)
diff$age[diff$diff < 6] <- "New"
diff$age[diff$diff >=6] <- "Old"
diff$age[is.na(diff$diff)] <- "Unknown"

diffplot = subset(diff, published_year > 2002 & published_year <2018, select=c(published_year, age))

# adding in a single record to get a point for 2005 - must be another way to do this...
extralev = data.frame(published_year = "2005", age = "Unknown")
diffplot = rbind(diffplot, extralev)

tax.nov.plot = subset(tax.nov.year, tax.nov.year$Year >="2003" & tax.nov.year$Year < "2018")

# now summarise the ratio of old and new for each red list assessment year
library(ggplot2)
fig_2 = ggplot() + 
  geom_bar(data = diffplot, aes(x = as.factor(published_year), fill = as.factor(age)), 
           position = position_dodge(preserve = 'single')) + 
  
  labs(x = "Red List Publication Year", y = "Number of plant species assessments", fill = "") +
  theme_minimal() +
  scale_fill_manual(values = c("New" = "#f37735", "Old" = "#ffc425", "Unknown" = "light grey")) +
  geom_point(data = tax.nov.plot, aes(x = as.factor(Year), y = Count), stat = "identity", size = 2, shape = 16) +
  #scale_colour_manual(name = "Year", labels = "test", values = "black") +
  scale_x_discrete(expand = c(0, 0)) + scale_y_continuous(expand = c(0,100)) 

fig_2

path = getwd()
#ggsave(paste0(path,"/03_figs/fig_2.png"), width = 10)
ggsave(paste0(path,"/fig_2.png"), width = 10)



```



### 3.3 Supporting the Plant assessment champions – Specialist Groups and Authorities

Compare number of species assessed vs not assessed for specialist groups and non-specialist groups

```{r}
#pull out just vascular plants
RL_vasc_plants = subset(all.plants, all.plants$phylum == "TRACHEOPHYTA" )
RL_vasc_plants$SG = ""

# total number of described vascualr plants 
described.vasc.plants.niclughadha2016 

# update SG with specialist group name
RL_vasc_plants$SG[RL_vasc_plants$family == "CYMODOCEACEAE"] = "Seagrass"
RL_vasc_plants$SG[RL_vasc_plants$family == "HYDROCHARITACEAE"] = "Seagrass"
RL_vasc_plants$SG[RL_vasc_plants$family == "POSIDONIACEAE"] = "Seagrass"
RL_vasc_plants$SG[RL_vasc_plants$family == "ZOSTERACEAE"] = "Seagrass"

RL_vasc_plants$SG[RL_vasc_plants$family == "ARECACEAE"] = "Palm"

RL_vasc_plants$SG[RL_vasc_plants$family == "ORCHIDACEAE"] = "Orchid"

RL_vasc_plants$SG[RL_vasc_plants$family == "CYCADACEAE"] = "Cycad"
RL_vasc_plants$SG[RL_vasc_plants$family == "ZAMIACEAE"] = "Cycad"
RL_vasc_plants$SG[RL_vasc_plants$family == "STANGERIACEAE"] = "Cycad"

RL_vasc_plants$SG[RL_vasc_plants$family == "ARAUCARIACEAE"] = "Conifer"
RL_vasc_plants$SG[RL_vasc_plants$family == "CUPRESSACEAE"] = "Conifer"
RL_vasc_plants$SG[RL_vasc_plants$family == "PINACEAE"] = "Conifer"
RL_vasc_plants$SG[RL_vasc_plants$family == "PODOCARPACEAE"] = "Conifer"
RL_vasc_plants$SG[RL_vasc_plants$family == "PHYLLOCLADACEAE"] = "Conifer"
RL_vasc_plants$SG[RL_vasc_plants$family == "SCIADOPITYACEAE"] = "Conifer"
RL_vasc_plants$SG[RL_vasc_plants$family == "TAXACEAE"] = "Conifer"
RL_vasc_plants$SG[RL_vasc_plants$family == "CEPHALOTAXACEAE"] = "Conifer"
RL_vasc_plants$SG[RL_vasc_plants$family == "STANGERIACEAE"] = "Conifer"

RL_vasc_plants$SG[RL_vasc_plants$family == "CACTACEAE"] = "Cacti & Succulents"

RL_vasc_plants$SG[RL_vasc_plants$family == "NEPENTHACEAE"] = "Carnivorous"
RL_vasc_plants$SG[RL_vasc_plants$family == "DROSERACEAE"] = "Carnivorous"
RL_vasc_plants$SG[RL_vasc_plants$family == "CEPHALOTACEAE"] = "Carnivorous"
RL_vasc_plants$SG[RL_vasc_plants$family == "SARRACENIACEAE"] = "Carnivorous"
RL_vasc_plants$SG[RL_vasc_plants$family == "LENTIBULARIACEAE"] = "Carnivorous"
RL_vasc_plants$SG[RL_vasc_plants$family == "RORIDULACEAE"] = "Carnivorous"
RL_vasc_plants$SG[RL_vasc_plants$family == "DIONCOPHYLLACEAE"] = "Carnivorous"

```

For each group (specialist group or not) get the count (n rows)
```{r}
#now group them up and get counts
library(tidyverse)
RL_vasc_noSG = subset(RL_vasc_plants, SG == "" )
Non_SG_assessed = nrow(RL_vasc_noSG)

RL_vasc_yesSG = subset(RL_vasc_plants, SG != "" )
SG_assessed = nrow(RL_vasc_yesSG)

```

Now import the count of species per family - derived from [Catalogue of life](http://www.catalogueoflife.org/col/browse/tree)
```{r}
path = getwd()
CoL_24thSep_2018 <- read.csv(paste0(path,"/01_data/CoL_24thSep_2018.csv"))

# get the total species count for SG groups
sumCoL = sum(CoL_24thSep_2018$Count)

# get the count of species in SGs not assessed
SGs_not_assessed = sumCoL - SG_assessed

# work out the remainder non SG and not assessed
Non_SG_not_assesseed = (described.vasc.plants.niclughadha2016 - sum(SG_assessed,SGs_not_assessed,Non_SG_assessed))

```

Combine to run the chi square test
```{r}
SGs = c(SG_assessed,SGs_not_assessed)
NonSGs = c(Non_SG_assessed,Non_SG_not_assesseed)

SG.test = as.data.frame(cbind(SGs,NonSGs))
names(SG.test) = c('assessed','not assessed')

rownames(SG.test) = c("SGs","NonSGs" )

chisq.test(SG.test)
```







